\chapter{FUTURE RESEARCH}
\label{chap:future}
There are many areas where the Pingo system can improve.  The first great improvement can happen in the evaluation methodology. All the workflows used to evaluate Pingo were probabilistically generated.   The generation system is designed with flexibility in mind so that it can generate different kinds of workflow loads to the system.  But there is no substitute to real data. Unfortunately, the amount of real data available to the researcher was not enough and was not completely relevant to the specific problem that Pingo attempts to solve. More real data needs to be gathered in order to produce more accurate evaluations on the performance of the decision algorithms of the system.

Another area of research is the implementation of more sophisticated adaptive decision algorithms that can handle the most diverse types of workloads submitted to the system.  In this respect, this area of research is very much interlinked to the previous proposition to gather a more diverse dataset of workflows. 

Most of the research done in the area of scientific workflows becomes relevant and applicable, not in the context of the \textbf{Decision System} of Pingo, but in the context of the metrics to evaluate the \textbf{Decision System}.  Most of the previous research has focused in finding optimal and near-optimal solutions to the problem of scientific workflows with constrained space, assuming that we know the entire history of the workflows that will be submitted to the system from the beginning.  In Pingo, the \textbf{Decision System} does not know the end from the beginning.  Instead, it only uses the previous workflows submitted to the system to predict how the future workflows might look like.  This different approach makes sense in fast-paced research settings where researchers don't know from the start the exact process (and hence the workflow) that they will follow in their research.

For evaluation purposes it is possible to know the end from the beginning and all the research from Chapter \ref{chap:related} becomes more directly relevant to our problem.  As a future endeavor, it would be good to do a more throughout exploratory work on the scientific workflows research literature in order to apply the most relevant produced results to the evaluation of the Pingo system.

Another important area of future research has to do with the design of the system. As we have seen, the system is nothing more than a composition of smaller independent subsystems that poll data from Hadoop or from a database that keeps the state of actions and datasets.  More research is needed on how to tune the parameters that control the frequency of this polling events, so that each independent subsystem carries its own processing computations as effectively as possible without putting to much strain in the underlying database cluster.

I am sure that the avid reader of this report will have identified some other opportunities in which the system can be improved or expanded. I gladly accept any related commentaries and suggestions about it. The most rewarding news for me as a researcher is that the system I have created is used and expanded and adapted to different needs by others. I certainly have attempted to design it with that goal in mind.